Fluent statistical computing interfaces for biological data analysis

### 2020-10-21 Rmd/05-embeddings.Rmd
typo corrections

While there has been much work on the algorithmic details of
embedding methods, there are relatively few tools designed to assist users
to interact with these techniques: when is an embedding sufficient for the
task at hand? Several
interactive interfaces have been proposed for evaluating or using embedding
techniques. @Buja2008-fn used tours to guide analysts during the optimisation
of multidimensional scaling methods by extending their interactive visualisation
software called **XGobi** and **GGobi** into a new tool called **GGvis** 
[@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
Their interface allows the analyst to dynamically modify and check whether
an MDS configuration has preserved the locality and closeness
of points between the configuration and the original data.
@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
non-linear embeddings in single cell RNA-seq data. It provides a click and 
highlight visualisation for colouring points in an embedding according to an 
estimated pairwise distance in the original high-dimensional space.
Similarly, the **TensorFlow**  embedding projector is a web interface to 
running some non-liner embedding methods live in the browser and 
provides interactions to colour points, and select nearest neighbours
[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
guided and modified form of the  t-SNE algorithm, that allows users to modify 
optimisation parameters in real-time.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. Instead, we implement a more pragmatic workflow by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks.

@@

To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produces $Y$ will remove noise in the original dataset while retaining any 
latent structure.

@@

$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
by the number of observations , $n-1$, in the data, corresponding to
$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].

@@

Like, when, using other DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections.  In practice, we first sphere 
our data via principal components to reduce dimensionality of $X$ prior to 
running the tour. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence of 'interesting' 
projections as quantified by an index function [@Cook1995-bi]. While our software, 
**liminal** is able to  visualise guided tours, our focus in the case studies 
uses the grand tour to see global structure in the data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near, or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

@@

From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

>>>>

While there has been much work on the algorithmic details of
embedding methods, there are relatively few tools designed to assist users
to interact with these techniques: when is an embedding sufficient for the
task at hand? Several
interactive interfaces have been proposed for evaluating or using embedding
techniques. @Buja2008-fn used tours to guide analysts during the optimisation
of multidimensional scaling methods by extending their interactive visualisation
software called **XGobi** and **GGobi** into a new tool called **GGvis** 
[@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
Their interface allows the analyst to dynamically modify and check whether
an MDS configuration has preserved the locality and closeness
of points between the configuration and the original data.
@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
non-linear embeddings in single cell RNA-seq data. It provides a click and 
highlight visualisation for colouring points in an embedding according to an 
estimated pairwise distance in the original high-dimensional space.
Similarly, the **TensorFlow**  embedding projector is a web interface to 
running some non-linear embedding methods live in the browser and 
provides interactions to colour points, and select nearest neighbours
[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
guided and modified form of the  t-SNE algorithm, that allows users to modify 
optimisation parameters in real-time.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are many parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. Instead, we implement a more pragmatic workflow by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks.

@@

To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produce $Y$ will remove noise in the original dataset while retaining any 
latent structure.

@@

$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
As the perplexity increases, $\sigma^2_{i}$ increases, until it is bounded above
by the number of observations , $n-1$, in the data, corresponding to
$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].

@@

Like when using other DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections.  In practice, we first sphere 
our data via principal components to reduce dimensionality of $X$ prior to 
running the tour. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence of 'interesting' 
projections as quantified by an index function [@Cook1995-bi]. While our software, 
**liminal** is able to  visualise guided tours, our focus in the case studies 
uses the grand tour to see global structure in the data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether points are near or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

@@

From the video linked in figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635905.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.
